---
title: "Global genomic epidemiology of multi-carbapenemase Enterobacterales "
subtitle: Code for plots and statistical analyses
keep-md: true
format: 
  html:
    toc: true
    embed-resources: true
author: 
  name: Nicole Lerminiaux
date: today
date-format: iso
---

# Setup
```{r}
#| label: setup
#| warning: false
library(dplyr)
library(tidyr)
library(stringr)
library(here)
library(readr)
library(ggplot2)
library(tibble)
library(patchwork)
library(ComplexUpset)
library(ggpattern)
library(gridpattern)
library(networkD3)
library(treemapify)
library(logistf)
library(Matrix)
library(rstatix)
library(goftest)

# load data
meta <- readRDS(here("data", "metadata.RDS"))
gene <- readRDS(here("data", "gene.RDS"))
contigs <- readRDS(here("data", "contig_list.RDS"))
amr <- readRDS(here("data", "amr.RDS"))
hvkp <- readRDS(here("data", "hvkp.RDS"))
ta <- readRDS(here("data", "toxin_antitoxin.RDS"))
meta_ncbi <- readRDS(here("data", "metadata_ncbi_enterobacterales.RDS"))
carbs_ncbi <- readRDS(here("data", "ncbi_enterobacterales_carbs.RDS"))
meta_raw <- readRDS(here("data", "metadata_raw.RDS"))
```

# Figures

### Figure 1: Most common carbapenemase groups coloured by genomic location
```{r}
#| label: fig-plot-1
#| fig-cap: Most common carbapenemase groups coloured by genomic location. 
#| fig-height: 4
#| warning: false

cupset_setup <- meta |> 
  separate_longer_delim(carb_group, delim = ", ") |> 
  select(accession, multicarb_location, genus, continent, isolation_source_group, carb_group) |> 
  distinct() |> 
  add_column(Present=1) |> 
  pivot_wider(names_from=carb_group, values_from=Present, values_fill = 0) |> 
  column_to_rownames("accession") |> 
  rename(`OXA-48-like` = OXA) |> 
  mutate(multicarb_location = factor(multicarb_location, levels = c("one plas", "two plas", "three or more plas", "one plas and chr", "two plas and chr", "three or more plas and chr", "chr")))
carb_list <- c("IMP", "KPC", "NDM", "OXA-48-like", "VIM")

upset_plt <- upset(cupset_setup, 
  carb_list,
  min_size = 5,
  name = "Carbapenemase combinations",
  base_annotations = list('Number of genomes' = intersection_size(mapping = aes(fill = multicarb_location), text = list(size = 3)) +
    scale_fill_manual(values = c("#ffc098", "#e9742a", "#93532a", "#f4c5ff", "#da62f5","#793a87", "#4182f3")) + 
    labs(fill = "Multi-carbapenemase\nlocation")),
  set_sizes=(upset_set_size() + 
    ylab('Count') + 
    theme(axis.text.x=element_text(angle=90, size = 8))),
  guides = "collect",
  width_ratio = 0.1);upset_plt
```

### Figure 2: Most common carbapenemase allelic combinations found across high-risk lineages

Note legend was coloured manually in Inkscape. 
```{r}
#| label: fig-plot-2
#| fig-cap: Most common carbapenemase allelic combinations found across high-risk lineages. 
#| fig-height: 6
#| warning: false

# set up stripe patterns
stripe_imp <- gridpattern::patternFill("stripe", fill = c("#D55E00", "#D55E00"), linewidth = NA, spacing = 0.1, density = 1)
stripe_imp_kpc <- gridpattern::patternFill("stripe", fill = c("#D55E00", "#56B4E9"), linewidth = NA, spacing = 0.1, density = 1)
stripe_imp_ndm <- gridpattern::patternFill("stripe", fill = c("#D55E00", "#E69F00"), linewidth = NA, spacing = 0.1, density = 1)
stripe_imp_oxa <- gridpattern::patternFill("stripe", fill = c("#D55E00", "#CC79A7"), linewidth = NA, spacing = 0.1, density = 1)
stripe_imp_vim <- gridpattern::patternFill("stripe", fill = c("#D55E00", "#0072B2"), linewidth = NA, spacing = 0.1, density = 1)
stripe_kpc <- gridpattern::patternFill("stripe", fill = c("#56B4E9", "#56B4E9"), linewidth = NA, spacing = 0.1, density = 1)
stripe_kpc_ndm <- gridpattern::patternFill("stripe", fill = c("#56B4E9", "#E69F00"), linewidth = NA, spacing = 0.1, density = 1)
stripe_kpc_oxa <- gridpattern::patternFill("stripe", fill = c("#56B4E9", "#CC79A7"), linewidth = NA, spacing = 0.1, density = 1)
stripe_kpc_oxa_vim <- gridpattern::patternFill("stripe", fill = c("#56B4E9", "#CC79A7", "#0072B2"), linewidth = NA, spacing = 0.1, density = 1)
stripe_kpc_vim <- gridpattern::patternFill("stripe", fill = c("#56B4E9", "#0072B2"), linewidth = NA, spacing = 0.1, density = 1)
stripe_ndm <- gridpattern::patternFill("stripe", fill = c("#E69F00", "#E69F00"), linewidth = NA, spacing = 0.1, density = 1)
stripe_ndm_oxa <- gridpattern::patternFill("stripe", fill = c("#E69F00", "#CC79A7"), linewidth = NA, spacing = 0.1, density = 1)
stripe_oxa <- gridpattern::patternFill("stripe", fill = c("#CC79A7", "#CC79A7"), linewidth = NA, spacing = 0.1, density = 1)
stripe_oxa_vim <- gridpattern::patternFill("stripe", fill = c("#CC79A7", "#0072B2"), linewidth = NA, spacing = 0.1, density = 1)
stripe_vim <- gridpattern::patternFill("stripe", fill = c("#0072B2", "#0072B2"), linewidth = NA, spacing = 0.1, density = 1)

meta |> 
  filter((genus == "Escherichia" & (mlst == "410" | mlst == "167" | mlst == "361")) |
  (str_detect(organism, "Klebsiella pneumoniae") & (mlst == "11" | mlst == "15" | mlst == "16" | mlst == "258" | mlst == "147" | mlst == "512")) | (genus == "Enterobacter" & (mlst == "171" | mlst == "79")) | (genus == "Citrobacter" & (mlst == "22"))) |> 
  count(genus, mlst, carb_group, carb) |> 
  mutate(carb_label = str_replace_all(carb, " ", "\n")) |> 
  mutate(carb_group = str_replace(carb_group, "OXA", "OXA-48-like")) |> 
  ggplot(aes(area=n, fill = carb_group)) +
  geom_treemap(color="white", show.legend = TRUE) + # fill = list(stripe_imp_ndm, stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc_ndm,stripe_kpc_oxa_vim,stripe_ndm_oxa,stripe_ndm_oxa,stripe_oxa,stripe_oxa,stripe_oxa_vim, stripe_vim,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc_ndm,stripe_ndm,stripe_ndm,stripe_ndm,stripe_ndm_oxa, stripe_ndm,stripe_ndm,stripe_ndm,stripe_ndm,stripe_ndm,stripe_ndm_oxa, stripe_ndm_oxa,stripe_ndm_oxa,stripe_kpc_ndm,stripe_kpc_ndm,stripe_ndm,stripe_ndm,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm,stripe_ndm,stripe_ndm,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_imp_kpc,stripe_imp_kpc,stripe_imp_oxa,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc_ndm,stripe_kpc_ndm,stripe_kpc_ndm,stripe_kpc_ndm,stripe_kpc_ndm,stripe_kpc_ndm,stripe_kpc_ndm,stripe_kpc_ndm,stripe_kpc_ndm,stripe_kpc_ndm,stripe_kpc_ndm,stripe_kpc_oxa,stripe_kpc_oxa,stripe_kpc_vim,stripe_ndm,stripe_ndm,stripe_ndm,stripe_ndm,stripe_ndm,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_oxa,stripe_imp_kpc,stripe_imp_vim,stripe_kpc_ndm,stripe_kpc_oxa,stripe_kpc_vim,stripe_ndm,stripe_ndm,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_oxa,stripe_oxa,stripe_oxa,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc_ndm,stripe_kpc_ndm,stripe_kpc_ndm,stripe_ndm,stripe_ndm,stripe_ndm,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_oxa,stripe_kpc,stripe_kpc_ndm,stripe_kpc_ndm,stripe_kpc_ndm,stripe_kpc_ndm,stripe_ndm,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_ndm_oxa,stripe_oxa,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc_ndm,stripe_kpc_vim,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc,stripe_kpc_ndm,stripe_kpc_oxa,stripe_kpc_oxa,stripe_kpc_oxa)) +
  geom_treemap_text(aes(label = paste0(n, "\n", carb_label)), place = "centre") +
  theme_bw() +
  facet_wrap(~ factor(mlst, levels = c("11", "16", "147", "258", "15", "512", "410", "167", "361", "171", "79", "22"), labels=c("K. pneumoniae\nST11\n(n=232)", "K. pneumoniae\nST16\n(n=62)","K. pneumoniae\nST147\n(n=57)","K. pneumoniae\nST258\n(n=60)","K. pneumoniae\nST15\n(n=39)","K. pneumoniae\nST512\n(n=31)","E. coli\nST410\n(n=30)","E. coli\nST167\n(n=15)","E. coli\nST361\n(n=13)","E. hormaechei\nST171\n(n=13)","E. hormaechei\nST79\n(n=13)","C. freundii\nST22\n(n=27)"))) +
  labs(fill = "Carbapenemase\ngroup")
# legend coloured manually in Inkscape
```

### Figure 4: Density of carbapenemase gene distances among proportion of plasmid length faceted by carbapenemase group

```{r}
#| label: fig-plot-4
#| fig-cap: Density of carbapenemase gene distances among proportion of plasmid length faceted by carbapenemase group. 
#| fig-height: 4
#| warning: false

tn2 <- gene |> 
  filter(location == "plasmid") |> 
  group_by(accession, contig_id, contig_length) |> 
  mutate(halfsize = contig_length/2) |>
  ungroup() 

plas_list <- gene |> filter(location == "plasmid") |> pull(contig_id) |> unique()
temp_list <- list()
for (i in 1:length(plas_list)){
  cnt <- tn2 |> filter(contig_id == plas_list[i]) |> nrow()
  
  temp <- tn2 |> filter(contig_id == plas_list[i])
  
  for (j in 1:cnt) {
    
    if(j < cnt){
    
    temp$diff2[j] <- ifelse((temp$start[j+1] - temp$start[j] < temp$halfsize[j]), 
            abs(temp$start[j+1] - temp$start[j]), 
            abs(temp$start[j+1] - temp$start[j] - temp$contig_length[j]))
           
    } else {
      
      temp$diff2[j] <-temp$start[1] + temp$contig_length[j] - temp$start[j]
    }
    
    temp_list[[i]] <- temp |> select(contig_id, start, diff2)
  }
}

tn3 <- tn2 |> left_join(bind_rows(temp_list), by = c("contig_id", "start"))

tn4 <- tn3 |> 
  filter(!(contig_length == diff2)) |> # remove those with only one carb on plas
  group_by(contig_id, contig_length, rep_type) |> 
  summarize(min = min(diff2)) |> 
  ungroup() |> 
  mutate(prop = min/contig_length) |> 
  left_join((gene |> group_by(contig_id) |> summarize(carb = toString(gene, sep = ", "))  |> filter(str_count(carb, ",") > 0)), by = "contig_id") |> 
  mutate(carb_group = case_when(
    str_detect(carb, "IMP") & str_detect(carb, "KPC") & str_detect(carb, "NDM") ~ "IMP, KPC, NDM",
    str_detect(carb, "IMP") & str_detect(carb, "KPC") & str_detect(carb, "OXA") ~ "IMP, KPC, OXA",
    str_detect(carb, "IMP") & str_detect(carb, "KPC") & str_detect(carb, "VIM") ~ "IMP, KPC, VIM",
    str_detect(carb, "IMP") & str_detect(carb, "NDM") & str_detect(carb, "OXA") ~ "IMP, NDM, OXA",
    str_detect(carb, "IMP") & str_detect(carb, "NDM") & str_detect(carb, "VIM") ~ "IMP, NDM, VIM",
    str_detect(carb, "IMP") & str_detect(carb, "OXA") & str_detect(carb, "VIM") ~ "IMP, OXA, VIM",
    str_detect(carb, "KPC") & str_detect(carb, "NDM") & str_detect(carb, "OXA") ~ "KPC, NDM, OXA",
    str_detect(carb, "KPC") & str_detect(carb, "NDM") & str_detect(carb, "VIM") ~ "KPC, NDM, VIM",
    str_detect(carb, "KPC") & str_detect(carb, "OXA") & str_detect(carb, "VIM") ~ "KPC, OXA, VIM",
    str_detect(carb, "NDM") & str_detect(carb, "OXA") & str_detect(carb, "VIM") ~ "NDM, OXA, VIM",
    str_detect(carb, "IMP") & str_detect(carb, "OXA") ~ "IMP, OXA",
    str_detect(carb, "IMP") & str_detect(carb, "NDM") ~ "IMP, NDM",
    str_detect(carb, "IMP") & str_detect(carb, "VIM") ~ "IMP, VIM",
    str_detect(carb, "IMP") & str_detect(carb, "KPC") ~ "IMP, KPC",
    str_detect(carb, "KPC") & str_detect(carb, "NDM") ~ "KPC, NDM",
    str_detect(carb, "KPC") & str_detect(carb, "OXA") ~ "KPC, OXA",
    str_detect(carb, "KPC") & str_detect(carb, "VIM") ~ "KPC, VIM",
    str_detect(carb, "NDM") & str_detect(carb, "OXA") ~ "NDM, OXA",
    str_detect(carb, "NDM") & str_detect(carb, "VIM") ~ "NDM, VIM",
    str_detect(carb, "OXA") & str_detect(carb, "VIM") ~ "OXA, VIM",
    str_detect(carb, "NDM") ~ "NDM",
    str_detect(carb, "OXA") ~ "OXA",
    str_detect(carb, "KPC") ~ "KPC",
    str_detect(carb, "IMP") ~ "IMP",
    str_detect(carb, "VIM") ~ "VIM",
    str_detect(carb, "IMI") ~ "IMI",
    TRUE ~ NA
  )) 

tn5 <- tn4 |> 
  mutate(carb_group = ifelse(carb_group != "NDM, OXA" & carb_group != "KPC, NDM" & carb_group != "KPC" & carb_group != "NDM" & carb_group != "OXA", "other", carb_group)) |> 
  mutate(carb_group = factor(carb_group, levels = c("KPC", "KPC, NDM", "NDM", "NDM, OXA", "OXA", "other"))) 

density_plt <- tn5 |>
  left_join(tn5 |> count(carb_group, name = "total_carb_group"), by = "carb_group") |> 
  mutate(label = paste0(carb_group, "\n(n=", total_carb_group, ")")) |> 
  mutate(label = str_replace(label, "OXA", "OXA-48-like")) |> 
  mutate(label = factor(label, levels = c("KPC\n(n=193)", "KPC, NDM\n(n=33)", "NDM\n(n=78)", "NDM, OXA-48-like\n(n=38)", "OXA-48-like\n(n=27)", "other\n(n=53)"))) |> 
  ggplot(aes(x = prop)) +
  geom_density(fill="#e9742a", color="#e9742a", alpha=0.5) +
  labs(y = "Density", x = "Plasmid length (proportion)") +
  facet_wrap(~ label, scales = "free_y") +
  theme_bw();density_plt
```


### Supplementary Figure 1: Proportion of all Enterobacterales genomes in NCBI (≤ 20 contigs) that encoded zero, one, or multiple carbapenemase genes

```{r}
#| label: fig-plot-s1
#| fig-cap: Proportion of all Enterobacterales genomes in NCBI (≤ 20 contigs) that encoded zero, one, or multiple carbapenemase genes. 
#| fig-height: 6
#| warning: false

ncbi_any_carb_accessions <- carbs_ncbi |> 
  distinct(accession) |> 
  pull()

ncbi_multicarb_accessions <- carbs_ncbi |> 
  count(accession)  |> 
  filter(n > 1) |> 
  pull(accession)

denominator <- meta_ncbi |> 
  filter(assembly_num_contigs <= 20)  |> 
  mutate(year = case_when(
    year == "-" ~ NA,
    year == "1016-08" ~ NA,
    str_detect(year, "-") ~ str_remove(year, "-.*"), 
    str_detect(year, "missing|Missing|unknown|N/A|n/a|na|not|Not|NOT|None|none|Unknown|restricted") ~ NA,
    str_detect(year, "/") ~ str_remove(year, ".*/"),
    TRUE ~ year)) |> 
  mutate(carb = ifelse(accession %in% ncbi_any_carb_accessions, "one", "none")) |> 
  mutate(carb = ifelse(accession %in% ncbi_multicarb_accessions, "multiple", carb)) |> 
  mutate(year = ifelse(year < 2000, "pre-2000", year)) |> 
  count(year, carb) |> 
  mutate(carb = factor(carb, levels = c("none", "one", "multiple"))) 

denom_a <- denominator |> 
  ggplot(aes(x = year, y = n, fill = carb)) +
  geom_bar(stat="identity") +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, vjust = 0.5)) +
  scale_fill_manual(values = c("cadetblue2", "cadetblue3", "cadetblue4")) +
  labs(y="Genome count\n(n=46,531 total)", x = "Year", fill="Carbapenemase\ngenes")

denom_b <- denominator |> 
  ggplot(aes(x = year, y = n, fill = carb)) +
  geom_bar(stat="identity",position="fill") +
  theme_bw() +
  theme(axis.text.x = element_text(angle=90, vjust = 0.5)) +
  scale_fill_manual(values = c("cadetblue2", "cadetblue3", "cadetblue4")) +
  labs(y="Genome proportion\n(n=46,531 total)", x = "Year", fill="Carbapenemase\ngenes")

denom_a + denom_b + plot_layout(nrow=2, guides = "collect") + plot_annotation(tag_level = "A")
```

### Supplementary Figure 2: Count of multi-carbapenemase combinations by continent over time

```{r}
#| label: fig-plot-s2
#| fig-cap: Count of multi-carbapenemase combinations by continent over time. 
#| fig-height: 4
#| warning: false

continents <- meta |> 
  #filter(dereplicate == "unique") |> 
  mutate(carb_group_plotting = case_when(
    str_count(carb_group, ",") > 1 ~ "other",
    carb_group == "NDM, VIM" ~ "other",
    carb_group == "IMP, OXA" ~ "other",
    carb_group == "IMP, VIM" ~ "other",
    TRUE ~ carb_group )) |> 
  mutate(carb_group_plotting = factor(carb_group_plotting, levels = c("IMP", "IMP, KPC", "IMP, NDM", "KPC", "KPC, NDM", "KPC, OXA", "KPC, VIM", "NDM", "NDM, OXA", "OXA", "OXA, VIM", "VIM", "other"))) |>     
  mutate(carb_group_plotting = str_replace(carb_group_plotting, "OXA", "OXA-48-like")) |> 
  filter(!is.na(year)) |> 
  filter(!is.na(continent)) |> 
  left_join(meta |> count(continent, name = "total_continent"), by = "continent") |>  
  mutate(label = paste0(continent, "\n(n=", total_continent, ")")) |> 
  group_by(year, continent, label, carb_group_plotting)  |> 
  count() |> 
  ggplot(aes(x = year, y = n, fill = carb_group_plotting)) +
  geom_bar_pattern(stat="identity", position = "stack", aes(pattern_fill = carb_group_plotting), pattern_colour = NA, pattern_density = 0.5, pattern_spacing = 0.02) + 
  theme_bw() +
  facet_wrap(~label)+
  scale_fill_manual(values = c("#D55E00", "#D55E00", "#D55E00", "#56B4E9", "#56B4E9", "#56B4E9", "#56B4E9", "#E69F00", "#E69F00", "#CC79A7", "#CC79A7", "#0072B2", "grey")) +
  scale_pattern_fill_manual(values = c("#D55E00", "#56B4E9", "#E69F00", "#56B4E9", "#E69F00", "#CC79A7", "#0072B2", "#E69F00", "#CC79A7", "#CC79A7", "#0072B2", "#0072B2", "grey")) +
  labs(fill = "Carbapenemase\ngroup", y = "Number of\ngenomes", x = "Year of collection", pattern_fill = "Carbapenemase\ngroup") +
  theme(axis.text.x = element_text(angle=90, vjust = 0.5, size = 6));continents
```

### Supplementary Figure 3: Linkage between carbapenemase allele combinations

```{r}
#| label: fig-plot-s3
#| fig-cap: Linkage between carbapenemase allele combinations. 
#| fig-height: 8
#| warning: false

doubles <- meta |> 
  #filter for samples that have only two carbs, exclude those with more
  filter(str_count(carb, ",") == 1) |> 
  mutate(carb = str_remove_all(carb, "bla")) |> 
  separate_wider_delim(carb, delim = ", ", names = c("carb1", "carb2"), too_few = "align_start") |> 
    mutate(carb1 = case_when(
    carb1 == "IMP" ~ "IMP novel",
    carb1 == "KPC" ~ "KPC novel",
    carb1 == "NDM" ~ "NDM novel",
    carb1 == "OXA" ~ "OXA-48-like novel",
    carb1 == "VIM" ~ "VIM novel",
    TRUE ~ carb1)) |> 
  mutate(carb2 = case_when(
    carb2 == "IMP" ~ "IMP novel",
    carb2 == "KPC" ~ "KPC novel",
    carb2 == "NDM" ~ "NDM novel",
    carb2 == "OXA" ~ "OXA-48-like novel",
    carb2 == "VIM" ~ "VIM novel",
    TRUE ~ carb2)) |> 
  select(source = carb1, target = carb2) 

# add unique names to each carb column so we don't get insane loopy plots by default
doubles$source <- paste0(doubles$source, "_q41")
doubles$target <- paste0(doubles$target, "_q42")

# filter out triples 
triples <- meta |> 
  filter(str_count(carb, ",") == 2) |> 
  mutate(carb = str_remove_all(carb, "bla")) |> 
  separate_wider_delim(carb, delim = ", ", names = c("carb1", "carb2", "carb3"), too_few = "align_start") |> 
  mutate(carb1 = case_when(
    carb1 == "IMP" ~ "IMP novel",
    carb1 == "KPC" ~ "KPC novel",
    carb1 == "NDM" ~ "NDM novel",
    carb1 == "OXA" ~ "OXA-48-like novel",
    carb1 == "VIM" ~ "VIM novel",
    TRUE ~ carb1)) |> 
  mutate(carb2 = case_when(
    carb2 == "IMP" ~ "IMP novel",
    carb2 == "KPC" ~ "KPC novel",
    carb2 == "NDM" ~ "NDM novel",
    carb2 == "OXA" ~ "OXA-48-like novel",
    carb2 == "VIM" ~ "VIM novel",
    TRUE ~ carb2)) |> 
  mutate(carb3 = case_when(
    carb3 == "IMP" ~ "IMP novel",
    carb3 == "KPC" ~ "KPC novel",
    carb3 == "NDM" ~ "NDM novel",
    carb3 == "OXA" ~ "OXA-48-like novel",
    carb3 == "VIM" ~ "VIM novel",
    TRUE ~ carb3)) |> 
  select(carb1, carb2, carb3)

# add unique names to each carb column so we don't get insane loopy plots by default
triples$carb1 <- paste0(triples$carb1, "_q41")
triples$carb2 <- paste0(triples$carb2, "_q42")
triples$carb3 <- paste0(triples$carb3, "_q43")

# make a new df that splits out carb2 and carb3 
temp <- triples |> 
  select(carb1 = carb2, carb2 = carb3)

# select carb1 and carb2 from original df, and bind to the carb2 and carb3 temp df. this capture first and second pair of every triple into two columns 
triples_condensed <- triples |> 
  select(carb1, carb2) |> 
  rbind(temp) |> 
  select(source = carb1, target = carb2) 

# combine the triples dataset with the doubles from above and count pairs 
all_condensed <- rbind(doubles, triples_condensed) |> 
  count(source, target, name = "value") |> 
  arrange(source, target)

# assign node numbers/row numbers and add group value for color
nodes <- data.frame(name = c(as.character(all_condensed$source), as.character(all_condensed$target)) |> unique()) |> 
  mutate(group = case_when(
    str_detect(name, "IMP") ~ "IMP",
    str_detect(name, "KPC") ~ "KPC",
    str_detect(name, "NDM") ~ "NDM",
    str_detect(name, "OXA") ~ "OXA",
    str_detect(name, "VIM") ~ "VIM",
    TRUE ~ NA)) 

# assign row number from nodes 
all_condensed$IDsource <- match(all_condensed$source, nodes$name)-1 
all_condensed$IDtarget <- match(all_condensed$target, nodes$name)-1

# remove gross suffix from plot
nodes$name <- sub('_q4[1-3]$', '', nodes$name)

# specify colors of bars
my_color <- 'd3.scaleOrdinal() .domain(["IMP", "KPC", "NDM", "OXA", "VIM"]) .range(["#D55E00", "#56B4E9", "#E69F00", "#CC79A7",  "#0072B2"])'

# plot
sankeyNetwork(Links = all_condensed, Nodes = nodes,
              Source = "IDsource", Target = "IDtarget",
              Value = "value", NodeID = "name",
              colourScale = my_color, NodeGroup="group", 
              iterations = 0,
              sinksRight=FALSE, fontSize = 12, nodeWidth = 30, fontFamily = "sans-serif")
```

### Supplementary Figure 4: Heatmap of carbapenemase allele combinations for genomes with two carbapenemase genes 

```{r}
#| label: fig-plot-s4
#| fig-cap: Heatmap of carbapenemase allele combinations (n=140/241) for genomes with two carbapenemase genes (86.0 %, 1290/1501). 
#| fig-height: 6
#| warning: false

combos <- meta |> 
  filter(str_count(carb, ",") == 1) |> 
  separate_wider_delim(carb, delim = ", ", names = c("carb1", "carb2")) |> 
  select(accession, carb1, carb2) |> 
  group_by(carb1, carb2) |> 
  count() |> 
  pivot_wider(id_cols = carb1, names_from = carb2, values_from = n) |> 
  pivot_longer(cols = !carb1, names_to = "carb2", values_to = "value") |> 
  ungroup() |> 
  mutate(carb1 = str_remove(carb1, "bla")) |> 
  mutate(carb2 = str_remove(carb2, "bla"))

# All possible values across BOTH columns
all_carbs <- sort(unique(c(combos$carb1, combos$carb2)))

combos3 <- tibble(carb1 = all_carbs, carb2 = all_carbs) |> 
  complete(carb1, carb2) |> 
  left_join(combos, by = c("carb1", "carb2")) |> 
  left_join((combos |> filter(carb1 != carb2)), by = c("carb1" = "carb2", "carb2" = "carb1")) |> 
  mutate_all(~replace(., is.na(.), 0)) |> 
  mutate(value = value.x + value.y) |> 
  select(carb1, carb2, value) |> 
  mutate(value = ifelse(value == 0, NA, value))  |> 
  mutate(carb1 = case_when(
    carb1 == "IMP" ~ "IMP novel",
    carb1 == "KPC" ~ "KPC novel",
    carb1 == "NDM" ~ "NDM novel",
    carb1 == "OXA" ~ "OXA-48-like novel",
    carb1 == "VIM" ~ "VIM novel",
    TRUE ~ carb1)) |> 
  mutate(carb2 = case_when(
    carb2 == "IMP" ~ "IMP novel",
    carb2 == "KPC" ~ "KPC novel",
    carb2 == "NDM" ~ "NDM novel",
    carb2 == "OXA" ~ "OXA-48-like novel",
    carb2 == "VIM" ~ "VIM novel",
    TRUE ~ carb2)) |> 
  mutate(carb1 = factor(carb1, levels = c("IMP-1","IMP-4","IMP-8","IMP-13","IMP-22","IMP-23","IMP-26","IMP-38","IMP-69","IMP-94","IMP-96","IMP novel","KPC-2","KPC-3","KPC-4","KPC-5","KPC-6","KPC-8","KPC-12","KPC-14","KPC-18","KPC-20","KPC-25","KPC-31","KPC-33","KPC-35","KPC-44","KPC-53","KPC-66","KPC-121","KPC-125","KPC-126","KPC-144","KPC-157","KPC-160","KPC-189","KPC-264","KPC novel","NDM-1","NDM-3","NDM-4","NDM-5","NDM-6","NDM-7","NDM-13","NDM-24","NDM novel","OXA-48","OXA-162","OXA-163","OXA-181","OXA-204","OXA-232","OXA-244","OXA-484","OXA-1181","OXA-1207","OXA-48-like novel","VIM-1","VIM-2","VIM-4","VIM-75","VIM-86","VIM novel"))) |> 
  mutate(carb2 = factor(carb2, levels = c("IMP-1","IMP-4","IMP-8","IMP-13","IMP-22","IMP-23","IMP-26","IMP-38","IMP-69","IMP-94","IMP-96","IMP novel","KPC-2","KPC-3","KPC-4","KPC-5","KPC-6","KPC-8","KPC-12","KPC-14","KPC-18","KPC-20","KPC-25","KPC-31","KPC-33","KPC-35","KPC-44","KPC-53","KPC-66","KPC-121","KPC-125","KPC-126","KPC-144","KPC-157","KPC-160","KPC-189","KPC-264","KPC novel","NDM-1","NDM-3","NDM-4","NDM-5","NDM-6","NDM-7","NDM-13","NDM-24","NDM novel","OXA-48","OXA-162","OXA-163","OXA-181","OXA-204","OXA-232","OXA-244","OXA-484","OXA-1181","OXA-1207","OXA-48-like novel","VIM-1","VIM-2","VIM-4","VIM-75","VIM-86","VIM novel")))  

heatmap <- combos3 |> 
  ggplot(aes(x = carb1, y = carb2, fill = value)) +
  geom_tile(color = "grey90") +
  geom_text(aes(label = ifelse(value > 0, value, "")), color = "white", size = 2) +
  theme_minimal()+
  theme(axis.text.x = element_text(angle = 90, vjust =0.5, hjust = 0.9, size=6),
    axis.text.y = element_text(size = 6)) +
  scale_fill_gradient(na.value="white", high="#264040",low="cadetblue3") +
  labs(fill = "Count", x = "Carbapenemase 1", y = "Carbapenemase 2");heatmap
```

### Supplementary Figure 5: Most common carbapenemase allelic combinations found across different genera

```{r}
#| label: fig-plot-s5
#| fig-cap: Most common carbapenemase allelic combinations found across different genera. 
#| fig-height: 5
#| warning: false

meta_carb <- meta  |> 
  #filter(dereplicate == "unique") |> 
  mutate(carb = ifelse(carb %in% (meta |> count(carb) |> filter(n > 9) |> pull(carb)), carb, "other")) |> 
  #mutate(carb = ifelse(carb %in% (meta |> filter(dereplicate == "unique") |> count(carb) |> filter(n > 9) |> pull(carb)), carb, "other")) |> 
  count(carb, genus)  |> 
  ungroup() |> 
  arrange(desc(n))  

genus <- meta_carb |> 
  mutate(carb = str_remove_all(carb, "bla")) |> 
  ggplot(aes(x=n, y = reorder(carb, n, FUN="sum"), fill = genus)) +
  geom_bar(stat="identity") +
  geom_text(data = (meta_carb |> mutate(carb = str_remove_all(carb, "bla")) |> group_by(carb)  |> summarize(n_genus = n_distinct(genus), n = sum(n), genus = "Klebsiella")), aes(y = carb, x = n, label = n_genus), hjust = -0.1, show.legend = FALSE, size = 3) +
  theme_bw() +
  labs(x = "Number of genomes", y = "Carbapenemase combinations\n(present in 10 or more genomes)", fill = "Genus") +
  theme(plot.margin = unit(c(0.5,0.5,0.5,0.5), "cm")) +
  scale_x_continuous(limits = c(0, 460)) +
  scale_fill_manual(values = c("#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C", "#FB9A99", "#E31A1C", "#FDBF6F", "#FF7F00", "#CAB2D6", "#6A3D9A", "#FFFF99", "#B15928"));genus
```

### Supplementary Figure 6: Genomes with three or more carbapenemases

```{r}
#| label: fig-plot-s6
#| fig-cap: Genomes with three or more carbapenemases. 
#| fig-height: 7
#| warning: false

tri_carb <- meta |> 
  filter(carb_count > 2)  |>  
  mutate(allele_description = case_when(
    multicarb_type == "same alleles" ~ "same alleles",
    carb == "blaKPC-2, blaKPC-2, blaNDM-1" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM-5, blaOXA-181, blaOXA-181" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC-2, blaNDM-1, blaNDM-1" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC, blaKPC-2, blaKPC-2" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM-5, blaNDM-5, blaOXA-181" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC, blaKPC-3, blaKPC-3" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC-2, blaKPC-2, blaNDM-5" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC-2, blaKPC-3, blaKPC-3" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC-2, blaNDM-1, blaOXA-48" ~ "distinct alleles",
    carb == "blaNDM-1, blaNDM-5, blaOXA-232" ~ "distinct alleles",
    carb == "blaKPC, blaKPC-2, blaKPC-2, blaKPC-25" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC-2, blaKPC-2, blaKPC-2, blaKPC-2, blaKPC-2, blaNDM-1" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC-2, blaKPC-2, blaNDM-1, blaNDM-1" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC-2, blaNDM-1, blaNDM-5" ~ "distinct alleles",
    carb == "blaKPC-2, blaNDM-1, blaVIM-24" ~ "distinct alleles",
    carb == "blaNDM, blaNDM-1, blaNDM-1" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM-1, blaOXA-232, blaOXA-232" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM-5, blaNDM-5, blaOXA-181, blaOXA-181" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM-5, blaNDM-5, blaOXA-48" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM-5, blaOXA-232, blaOXA-232" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaIMP-11, blaKPC-2, blaNDM-5" ~ "distinct alleles",
    carb == "blaIMP-4, blaKPC-2, blaKPC-2" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaIMP-4, blaKPC-2, blaNDM-1, blaNDM-4" ~ "distinct alleles",
    carb == "blaIMP-4, blaKPC-2, blaNDM-5" ~ "distinct alleles",
    carb == "blaIMP-4, blaKPC-3, blaNDM-1" ~ "distinct alleles",
    carb == "blaIMP-4, blaNDM-1, blaNDM-1" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaIMP-8, blaKPC-2, blaNDM-1" ~ "distinct alleles",
    carb == "blaKPC, blaKPC, blaKPC-3" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC, blaKPC, blaNDM-1" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC, blaKPC, blaVIM-4, blaVIM-4" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC, blaKPC-2, blaNDM-1" ~ "distinct alleles",
    carb == "blaKPC, blaKPC-2, blaVIM-1" ~ "distinct alleles",
    carb == "blaKPC, blaKPC-3, blaKPC-3, blaKPC-3" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC-2, blaKPC-2, blaKPC-2, blaKPC-2, blaNDM-1" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC-2, blaNDM, blaNDM" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC-2, blaNDM, blaNDM, blaNDM" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC-2, blaNDM, blaNDM-1, blaNDM-1" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC-2, blaNDM-5, blaNDM-5, blaNDM-5, blaNDM-5" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC-2, blaOXA-181, blaVIM-1" ~ "distinct alleles",
    carb == "blaKPC-2, blaVIM, blaVIM-1" ~ "distinct alleles",
    carb == "blaKPC-3, blaNDM-5, blaNDM-5, blaNDM-5, blaNDM-5" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaKPC-3, blaNDM-7, blaNDM-7" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM, blaNDM, blaNDM-1" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM, blaNDM, blaNDM-1, blaNDM-1" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM, blaNDM, blaNDM-5" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM, blaNDM, blaNDM-5, blaNDM-5, blaNDM-5, blaNDM-5" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM, blaNDM-5, blaNDM-5" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM, blaNDM-5, blaNDM-5, blaNDM-5" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM, blaOXA-232, blaOXA-232, blaOXA-232" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM-1, blaNDM-1, blaNDM-7" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM-1, blaNDM-1, blaOXA-181" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM-1, blaNDM-1, blaOXA-48" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM-1, blaNDM-4, blaOXA-181" ~ "distinct alleles",
    carb == "blaNDM-1, blaNDM-4, blaOXA-232" ~ "distinct alleles",
    carb == "blaNDM-1, blaNDM-5, blaOXA-181" ~ "distinct alleles",
    carb == "blaNDM-1, blaNDM-5, blaOXA-232, blaOXA-232" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM-1, blaOXA, blaOXA-181" ~ "distinct alleles",
    carb == "blaNDM-1, blaOXA-232, blaOXA-48" ~ "distinct alleles",
    carb == "blaNDM-1, blaOXA-48, blaOXA-48" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM-1, blaOXA-48, blaVIM-1" ~ "distinct alleles",
    carb == "blaNDM-5, blaNDM-5, blaNDM-5, blaOXA-181" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaNDM-5, blaOXA, blaOXA-232" ~ "distinct alleles",
    carb == "blaNDM-5, blaOXA-181, blaOXA-48" ~ "distinct alleles",
    carb == "blaOXA, blaOXA, blaOXA-232, blaOXA-232" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaOXA-48, blaOXA-48, blaVIM-63, blaVIM-63, blaVIM-63, blaVIM-63, blaVIM-63" ~ "at least one duplicated\nallele plus other gene(s)",
    carb == "blaVIM-1, blaVIM-2, blaVIM-2, blaVIM-2" ~ "at least one duplicated\nallele plus other gene(s)",
    TRUE ~ NA
  )) 

tri_carb_loc <- tri_carb |> 
  mutate(allele_description = factor(allele_description, levels = c("same alleles", "at least one duplicated\nallele plus other gene(s)", "distinct alleles"))) |> 
  mutate(multicarb_location = factor(multicarb_location, levels = c("one plas", "two plas", "three or more plas", "one plas and chr", "two plas and chr", "three or more plas and chr", "chr"))) |> 
  ggplot(aes(x = allele_description, fill = multicarb_location)) + 
  geom_bar(stat = "count") +
  theme_bw() + 
  labs(y = "Number of genomes", x = "Allele types for genomes with\nthree or more carbapenemases", fill = "Multicarbapenemase\nlocation") +
  scale_fill_manual(values = c("#ffc098", "#e9742a",   "#93532a", "#f4c5ff", "#da62f5","#793a87", "#4182f3"))

tri_carb_counts <- tri_carb  |> 
  mutate(carb = ifelse(carb %in% (tri_carb |> count(carb) |> filter(n > 2) |> pull(carb)), carb, "other")) |> 
  count(carb, genus)  |> 
  ungroup() |> 
  arrange(desc(n))  

tri_carb_gen <- tri_carb_counts |>
  mutate(carb = str_remove_all(carb, "bla")) |> 
  ggplot(aes(x=n, y = reorder(carb, n, FUN="sum"), fill = genus)) +
  geom_bar(stat="identity") +
  geom_text(data = (tri_carb_counts |> mutate(carb = str_remove_all(carb, "bla")) |> group_by(carb)  |> summarize(n_genus = n_distinct(genus), n = sum(n), genus = "Klebsiella")), aes(y = carb, x = n, label = n_genus), hjust = -0.4, show.legend = FALSE, size = 3) +
  theme_bw() +
  labs(x = "Number of genomes", y = "Carbapenemase combinations\n(present in 3 or more genomes)", fill = "Genus") +
  scale_fill_manual(values = c("#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C", "#FB9A99", "#E31A1C", "#FDBF6F", "#FF7F00", "#CAB2D6", "#6A3D9A", "#FFFF99", "#B15928"))

tri_carb_gen + free(tri_carb_loc, side = "l") + guide_area() + plot_layout(design = "AC\nBC", guides = "collect", heights = c(0.65, 0.45)) + plot_annotation(tag_level="A")
```

### Supplementary Figure 7: Proportion of AMR-encoding plasmids that encoded at least one resistance gene to different antibiotic gene classes

```{r}
#| label: fig-plot-s7
#| fig-cap: Proportion of AMR-encoding plasmids that encoded at least one resistance gene to different antibiotic gene classes. 
#| fig-height: 4
#| warning: false

multicarb_plas_list <- gene |> 
  filter(location == "plasmid")  |> 
  count(contig_id) |> 
  filter(n > 1) |> 
  pull(contig_id)

singlecarb_plas_list <- gene |> 
  filter(location == "plasmid")  |> 
  count(contig_id) |> 
  filter(n == 1) |> 
  pull(contig_id)

amr_nocarbs <- amr |> 
  filter(!str_detect(gene, "blaNDM|blaKPC|blaOXA-48|blaOXA-181|blaOXA-232|blaOXA-244|blaOXA-204|blaOXA-1205|blaOXA-1207|blaVIM|blaIMP") & !str_detect(closest_reference, "OXA-48|OXA-181|OXA-232|OXA-244|OXA-204|OXA-1205|OXA-1207")) 

multicarb_plas_amr1 <- gene |> 
  filter(location == "plasmid") |> 
  mutate(carb_count = ifelse(contig_id %in% multicarb_plas_list, "Two or more", ifelse(contig_id %in% singlecarb_plas_list, "One", "None"))) |> 
  group_by(carb_count, contig_id) |> 
  arrange(gene) |> 
  summarize(carb = toString(gene, sep = ", ")) |> 
  mutate(carb_group = case_when(
    str_detect(carb, "IMP") & str_detect(carb, "KPC") & str_detect(carb, "NDM") ~ "IMP, KPC, NDM",
    str_detect(carb, "IMP") & str_detect(carb, "KPC") & str_detect(carb, "OXA") ~ "IMP, KPC, OXA",
    str_detect(carb, "IMP") & str_detect(carb, "KPC") & str_detect(carb, "VIM") ~ "IMP, KPC, VIM",
    str_detect(carb, "IMP") & str_detect(carb, "NDM") & str_detect(carb, "OXA") ~ "IMP, NDM, OXA",
    str_detect(carb, "IMP") & str_detect(carb, "NDM") & str_detect(carb, "VIM") ~ "IMP, NDM, VIM",
    str_detect(carb, "IMP") & str_detect(carb, "OXA") & str_detect(carb, "VIM") ~ "IMP, OXA, VIM",
    str_detect(carb, "KPC") & str_detect(carb, "NDM") & str_detect(carb, "OXA") ~ "KPC, NDM, OXA",
    str_detect(carb, "KPC") & str_detect(carb, "NDM") & str_detect(carb, "VIM") ~ "KPC, NDM, VIM",
    str_detect(carb, "KPC") & str_detect(carb, "OXA") & str_detect(carb, "VIM") ~ "KPC, OXA, VIM",
    str_detect(carb, "NDM") & str_detect(carb, "OXA") & str_detect(carb, "VIM") ~ "NDM, OXA, VIM",
    str_detect(carb, "IMP") & str_detect(carb, "OXA") ~ "IMP, OXA",
    str_detect(carb, "IMP") & str_detect(carb, "NDM") ~ "IMP, NDM",
    str_detect(carb, "IMP") & str_detect(carb, "VIM") ~ "IMP, VIM",
    str_detect(carb, "IMP") & str_detect(carb, "KPC") ~ "IMP, KPC",
    str_detect(carb, "KPC") & str_detect(carb, "NDM") ~ "KPC, NDM",
    str_detect(carb, "KPC") & str_detect(carb, "OXA") ~ "KPC, OXA",
    str_detect(carb, "KPC") & str_detect(carb, "VIM") ~ "KPC, VIM",
    str_detect(carb, "NDM") & str_detect(carb, "OXA") ~ "NDM, OXA",
    str_detect(carb, "NDM") & str_detect(carb, "VIM") ~ "NDM, VIM",
    str_detect(carb, "OXA") & str_detect(carb, "VIM") ~ "OXA, VIM",
    str_detect(carb, "NDM") ~ "NDM",
    str_detect(carb, "OXA") ~ "OXA",
    str_detect(carb, "KPC") ~ "KPC",
    str_detect(carb, "IMP") ~ "IMP",
    str_detect(carb, "VIM") ~ "VIM",
    str_detect(carb, "IMI") ~ "IMI",
    TRUE ~ NA
  )) |> 
  mutate(carb_group = ifelse(carb_group != "NDM, OXA" & carb_group != "KPC, NDM" & carb_group != "KPC" & carb_group != "NDM" & carb_group != "OXA" & carb_count == "Two or more", "other", carb_group)) 

nocarb_plas_amr <- contigs |> 
  filter(carb == "no" & location == "plasmid") |> 
  select(carb_count = carb, contig_id) |> 
  mutate(carb = NA, carb_group = "None", carb_count = "None")   

multicarb_plas_amr2 <- rbind(multicarb_plas_amr1, nocarb_plas_amr) |> 
  left_join((amr_nocarbs |> select(contig_id, accession, gene, subclass)), by = c("contig_id")) |> 
  separate_longer_delim(subclass, delim = "/") |> 
  filter(!(carb_count == "None" & is.na(gene))) |> 
  select(!gene) |> 
  distinct() |> 
  pivot_wider(id_cols = c("contig_id", "carb_count", "carb", "carb_group", "accession"), names_from = subclass, values_from = subclass, values_fn = ~1, values_fill = 0) |> 
  pivot_longer(cols = !c("contig_id", "carb_count",  "carb", "carb_group", "accession"), names_to = "antibiotic", values_to = "n") |> 
  filter(antibiotic != "NA") 

num_zero_carb_plas <- multicarb_plas_amr2 |> 
  distinct(contig_id, carb_count)  |> 
  count(carb_count) |> 
  filter(carb_count == "None") |> 
  pull(n)

amr_heatmap <- multicarb_plas_amr2 |> 
  group_by(carb_count, carb_group, antibiotic) |>
  summarize(sum = sum(n)) |> 
  left_join((multicarb_plas_amr1  |> count(carb_count, carb_group, name = "total_plas")), by = c("carb_group", "carb_count")) |> 
  mutate(total_plas = ifelse(carb_count == "None", num_zero_carb_plas, total_plas)) |> 
  mutate(prop = sum/total_plas) |> 
  mutate(carb_group = str_replace(carb_group, "OXA", "OXA-48-like")) |> 
  mutate(label = paste0(carb_group, "\n(n=", total_plas, ")" )) |> 
  ggplot(aes(x = label, y = antibiotic)) +
   geom_tile(aes(fill = prop)) +
  geom_text(aes(label = round(prop, digits=2)), color="white", size=3) +
   theme_bw() +
   theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 0.95)) +
  facet_wrap(~ carb_count, scales = "free_x", space = "free_x") +
  scale_fill_gradient(low="#264040",high="cadetblue3") +
  labs(x = "Carbapenemase group", y = "Antibiotic class", fill = "Proportion\nof plasmids\nencoding\nat least\none gene");amr_heatmap
```

### Supplementary Figure 8: Carbapenemase groups and genera represented in the top 8 carbapenemase plasmid secondary clusters

```{r}
#| label: fig-plot-s8
#| fig-cap: Carbapenemase groups and genera represented in the top 8 carbapenemase plasmid secondary clusters. 
#| fig-height: 6
#| warning: false

gene_carb_group <- gene |> 
  filter(location == "plasmid") |> 
  filter(str_detect(secondary_cluster, "AH615|AH539|AL194|AH594|AJ272|AH561|AI283|AI436")) |> 
  group_by(contig_id, organism, primary_cluster, secondary_cluster) |> 
  arrange(gene) |> 
  summarize(carb = toString(gene, sep =", ")) |> 
  ungroup() |> 
  mutate(carb_group = case_when(
    str_detect(carb, "IMP") & str_detect(carb, "KPC") & str_detect(carb, "NDM") ~ "IMP, KPC, NDM",
    str_detect(carb, "IMP") & str_detect(carb, "KPC") & str_detect(carb, "OXA") ~ "IMP, KPC, OXA",
    str_detect(carb, "IMP") & str_detect(carb, "KPC") & str_detect(carb, "VIM") ~ "IMP, KPC, VIM",
    str_detect(carb, "IMP") & str_detect(carb, "NDM") & str_detect(carb, "OXA") ~ "IMP, NDM, OXA",
    str_detect(carb, "IMP") & str_detect(carb, "NDM") & str_detect(carb, "VIM") ~ "IMP, NDM, VIM",
    str_detect(carb, "IMP") & str_detect(carb, "OXA") & str_detect(carb, "VIM") ~ "IMP, OXA, VIM",
    str_detect(carb, "KPC") & str_detect(carb, "NDM") & str_detect(carb, "OXA") ~ "KPC, NDM, OXA",
    str_detect(carb, "KPC") & str_detect(carb, "NDM") & str_detect(carb, "VIM") ~ "KPC, NDM, VIM",
    str_detect(carb, "KPC") & str_detect(carb, "OXA") & str_detect(carb, "VIM") ~ "KPC, OXA, VIM",
    str_detect(carb, "NDM") & str_detect(carb, "OXA") & str_detect(carb, "VIM") ~ "NDM, OXA, VIM",
    str_detect(carb, "IMP") & str_detect(carb, "OXA") ~ "IMP, OXA",
    str_detect(carb, "IMP") & str_detect(carb, "NDM") ~ "IMP, NDM",
    str_detect(carb, "IMP") & str_detect(carb, "VIM") ~ "IMP, VIM",
    str_detect(carb, "IMP") & str_detect(carb, "KPC") ~ "IMP, KPC",
    str_detect(carb, "KPC") & str_detect(carb, "NDM") ~ "KPC, NDM",
    str_detect(carb, "KPC") & str_detect(carb, "OXA") ~ "KPC, OXA",
    str_detect(carb, "KPC") & str_detect(carb, "VIM") ~ "KPC, VIM",
    str_detect(carb, "NDM") & str_detect(carb, "OXA") ~ "NDM, OXA",
    str_detect(carb, "NDM") & str_detect(carb, "VIM") ~ "NDM, VIM",
    str_detect(carb, "OXA") & str_detect(carb, "VIM") ~ "OXA, VIM",
    str_detect(carb, "NDM") ~ "NDM",
    str_detect(carb, "OXA") ~ "OXA",
    str_detect(carb, "KPC") ~ "KPC",
    str_detect(carb, "IMP") ~ "IMP",
    str_detect(carb, "VIM") ~ "VIM",
    str_detect(carb, "IMI") ~ "IMI",
    TRUE ~ NA
  ))
  
cluster_a <- gene_carb_group |> 
  mutate(carb_group_plotting = carb_group ) |> 
  mutate(carb_group_plotting = factor(carb_group_plotting, levels = c("IMP", "IMP, KPC", "IMP, VIM", "KPC", "KPC, NDM", "KPC, VIM", "NDM", "NDM, OXA", "OXA", "VIM"))) |>     
  group_by(primary_cluster, secondary_cluster, carb_group_plotting)  |> 
  count() |> 
  mutate(cluster = paste0(primary_cluster, "/\n", secondary_cluster)) |> 
  mutate(carb_group_plotting = str_replace(carb_group_plotting, "OXA", "OXA-48-like")) |> 
  ggplot(aes(x = cluster, y = n, fill = carb_group_plotting)) +
  geom_bar_pattern(stat="identity", position = "stack", aes(pattern_fill = carb_group_plotting), pattern_colour = NA, pattern_density = 0.5, pattern_spacing = 0.02) + 
  theme_bw() +
  theme(axis.text.x = element_text(size = 7, angle=90, vjust = 0.5)) +
  scale_fill_manual(values = c("#D55E00", "#D55E00", "#56B4E9", "#56B4E9", "#56B4E9", "#E69F00", "#E69F00", "#CC79A7", "#0072B2")) +
  scale_pattern_fill_manual(values = c("#56B4E9", "#0072B2", "#56B4E9", "#E69F00", "#0072B2", "#E69F00", "#CC79A7", "#CC79A7", "#0072B2")) +
  labs(fill = "Carbapenemase\ngroup", y = "Number of\nplasmids", x = "Plasmid cluster", pattern_fill = "Carbapenemase\ngroup")

cluster_b <- gene_carb_group |> 
  mutate(genus = str_remove(organism, " .*")) |> 
  group_by(primary_cluster, secondary_cluster, genus)  |> 
  count() |> 
  mutate(cluster = paste0(primary_cluster, "/\n", secondary_cluster)) |> 
  ggplot(aes(x = cluster, y = n, fill = genus)) +
  geom_bar(stat="identity", position="stack") +
  theme_bw() +
  theme(axis.text.x = element_text(size = 7, angle=90, vjust = 0.5)) +
  scale_fill_manual(values = c("#A6CEE3", "#1F78B4", "#B2DF8A", "#33A02C", "#FDBF6F", "#CAB2D6", "#6A3D9A")) +
  labs(fill = "Genus", y = "Number of\nplasmids", x = "Plasmid cluster")
  
cluster_a + cluster_b + plot_layout(nrow = 2) + plot_annotation(tag_level ="A") & theme(legend.justification = "left")
```

### Supplementary Figure 9: Hypervirulent *K. pneumoniae*

```{r}
#| label: fig-plot-s9
#| fig-cap: Hypervirulent *K. pneumoniae*. 
#| fig-height: 7
#| warning: false

hvkp_meta <- hvkp |> 
  ungroup() |> 
  distinct(accession) |> 
  left_join(meta |> select(accession, mlst, year, country, continent, isolation_source_group, carb, carb_group, carb_count, multicarb_location, multicarb_type)) 

hvkp_combos <- hvkp_meta |> 
  group_by(carb_group) |> 
  count(carb) |> 
  mutate(carb = ifelse(n == 1, "other", carb)) |> 
  group_by(carb_group, carb) |> 
  mutate(carb = str_remove_all(carb, "bla")) |> 
  mutate(carb = case_when(
    carb == "OXA, OXA" ~ "OXA novel, OXA novel", 
    carb == "KPC, KPC-2" ~ "KPC novel, KPC-2", 
    carb == "OXA, OXA-232" ~ "OXA novel, OXA-232", 
    carb == "KPC, KPC-2, KPC-2, KPC-25" ~ "KPC novel, KPC-2, KPC-2, KPC-25", 
    TRUE ~ carb)) |> 
  summarize(nn = sum(n)) |> 
  mutate(carb_group = str_replace(carb_group, "OXA", "OXA-48-like")) |> 
  ggplot(aes(x=nn, y = reorder(carb,nn, FUN = "sum"), fill = carb_group)) +
  geom_bar_pattern(stat="identity", position = "stack", aes(pattern_fill = carb_group), pattern_colour = NA, pattern_density = 0.5, pattern_spacing = 0.02) + 
  theme_bw() +
  scale_fill_manual(values = c("#D55E00", "#D55E00", "#56B4E9", "#56B4E9", "#56B4E9", "#56B4E9","#56B4E9","#E69F00", "#E69F00", "#CC79A7")) +
  scale_pattern_fill_manual(values = c("#56B4E9", "#E69F00", "#56B4E9", "#E69F00", "grey30","#CC79A7","#0072B2", "#E69F00", "#CC79A7", "#CC79A7")) +
  labs(fill = "Carbapenemase\ngroup", y = "Carbapenemase combinations\n(present in 2 or more genomes)", x = "Number of genomes", pattern_fill = "Carbapenemase\ngroup")

hvkp_plas <- hvkp |> 
  left_join(contigs |> select(contig_id, secondary_cluster, length, location), by = "contig_id") |> 
  filter(location == "plasmid") |> 
  mutate(carb = ifelse(contig_id %in% (gene |> distinct(contig_id) |> pull()), "yes", "no")) |> 
  ungroup() 

multicarb_location_hvkp <- gene |> 
  filter(accession %in% (hvkp_meta |> pull(accession))) |> 
  mutate(genomic_location = ifelse(location == "plasmid" & contig_id %in% (hvkp_plas |> distinct(contig_id) |> pull(contig_id)), "virplasmid", location)) |> 
  mutate(contig_location = paste(contig_id, genomic_location, sep = "_")) |> 
  group_by(accession, contig_location) |> 
  summarize(gene= toString(gene, sep= ", ")) |> 
  summarize(contig_location = toString(contig_location, sep = ", ")) |> 
  mutate(multicarb_location = case_when(
    str_detect(contig_location, "_plasmid, .*_plasmid.*_plasmid") & str_detect(contig_location, "chromosome") ~ "three or more non-vir plas and chr",
    str_detect(contig_location, "_plasmid, .*_plasmid") & str_detect(contig_location, "chromosome") ~ "two non-vir plas and chr",
    str_detect(contig_location, "_virplasmid, .*plasmid") & str_detect(contig_location, "chromosome") ~ "vir plas, non-vir plas, and chr",
    str_detect(contig_location, "_plasmid") & str_detect(contig_location, "chromosome") ~ "non-vir plas and chr",
    str_detect(contig_location, "_virplasmid") & str_detect(contig_location, "chromosome") ~ "vir plas and chr",
    str_detect(contig_location, "_plasmid, .*_plasmid, .*_plasmid") ~ "three or more non-vir plas",
    str_detect(contig_location, "_virplasmid, .*_plasmid, .*_plasmid") ~ "vir plas and two or more non-vir plas",
    str_detect(contig_location, "_plasmid, .*_virplasmid, .*_plasmid") ~ "vir plas and two or more non-vir plas",
    str_detect(contig_location, "_plasmid, .*_plasmid, .*_virplasmid") ~ "vir plas and two or more non-vir plas",
    str_detect(contig_location, "_plasmid, .*_plasmid") ~ "two non-vir plas",
    str_detect(contig_location, "_virplasmid") & str_detect(contig_location, "_plasmid") ~ "vir plas and non-vir plas",
    str_detect(contig_location, "chromosome")  ~ "chr",
    str_detect(contig_location, "_plasmid") ~ "non-vir plas",
    str_detect(contig_location, "_virplasmid") ~ "vir plas",
    TRUE ~ NA  
  )) |> 
  select(!contig_location)

hvkp_location <- hvkp_meta |> 
  left_join(multicarb_location_hvkp, by = "accession") |> 
  mutate(multicarb_location.y = factor(multicarb_location.y, levels = c("non-vir plas", "vir plas",  "two non-vir plas", "vir plas and non-vir plas",  "three or more non-vir plas", "vir plas and two or more non-vir plas",  "non-vir plas and chr", "vir plas and chr",  "two non-vir plas and chr", "vir plas, non-vir plas, and chr", "three or more non-vir plas and chr", "chr"))) |> 
  mutate(multicarb_type = ifelse(multicarb_type == "same alleles", "same\nalleles", ifelse(multicarb_type == "different alleles", "different\nalleles", "different\nfamilies"))) |> 
  mutate(multicarb_type = factor(multicarb_type, levels = c("same\nalleles", "different\nalleles", "different\nfamilies"))) |> 
  ggplot(aes(x = multicarb_type, fill = multicarb_location.y)) + 
  geom_bar_pattern(stat="count", position = "stack", aes(pattern_fill = multicarb_location.y), pattern_colour = NA, pattern_density = 0.5, pattern_spacing = 0.02) + 
  theme_bw() +
  scale_fill_manual(values = c("#ffc098" , "#ffc098", "#e9742a", "#e9742a", "#93532a", "#93532a", "#f4c5ff" , "#f4c5ff", "#da62f5", "#da62f5", "#793a87", "#4182f3")) +
  scale_pattern_fill_manual(values = c("#ffc098" , "grey30", "#e9742a", "grey30", "#93532a", "grey30", "#f4c5ff" , "grey30", "#da62f5", "grey30", "#793a87", "#4182f3")) +
  labs(y = "Number of genomes", x = "Multi-carbapenemase type", fill = "Multi-carbapenemase\nlocation", pattern_fill = "Multi-carbapenemase\nlocation")

hvkp_combos + free(hvkp_location, side="l") + plot_layout(design = "A\nB", heights =c(0.65, 0.4)) + plot_annotation(tag_levels = "A") & theme(legend.justification = "left")
```

### Supplementary Figure 10: Carbapenemases with non-exact allele matches 

```{r}
#| label: fig-plot-s10
#| fig-cap: Carbapenemases with non-exact allele matches. 
#| fig-height: 4
#| warning: false

mismatches2 <- amr |> 
  filter(accession %in% (meta |> pull(accession))) |> 
  filter(str_detect(gene, "blaNDM|blaKPC|blaOXA-48|blaOXA-181|blaOXA-232|blaOXA-244|blaOXA-204|blaOXA-1205|blaOXA-1207|blaVIM|blaIMP") | str_detect(closest_reference, "OXA-48|OXA-181|OXA-232|OXA-244|OXA-204|OXA-1205|OXA-1207")) |> 
  filter(status != "ALLELEX" & status != "PARTIAL_CONTIG_ENDX") |> 
  left_join(meta_raw |> select(accession, multicarb_type, multicarb_location, carb), by = "accession") |> 
  filter(!is.na(multicarb_type)) |> 
  mutate(multicarb_type = ifelse(multicarb_type == "different types", "different families", "same family")) |> 
  group_by(gene, status) |> 
  count(multicarb_type)

mismatches_plt <- mismatches2 |> 
  mutate(multicarb_type = factor(multicarb_type, levels = c("same family", "different families"))) |> 
  mutate(status = case_when(
    status == "BLASTX" ~ "different protein",
    status == "PARTIALX" ~ "pseudogene/frameshift",
    status == "INTERNAL_STOP" ~ "internal stop",
    #status == "PARTIAL_CONTIG_ENDX" ~ "split on contig end",
    TRUE ~ NA)) |> 
  ggplot(aes(x = multicarb_type, fill = status, y = n)) +
  geom_bar(stat = "identity") +
  theme_bw() +
  scale_fill_manual(values = c("cadetblue2", "cadetblue3", "cadetblue4")) +
  labs(x = "Carbapenemases in genome", y = "Mutant carbapenemase\n(non-exact allele matches)\ncount", fill = "Gene status") +
  theme(axis.text.x = element_text(angle=90)) +
  facet_wrap(~ factor(gene, labels = c("IMP\n(n=1)", "KPC\n(n=75)", "NDM\n(n=46)", "OXA-48-\nlike\n(n=22)", "VIM\n(n=3)")), nrow=1);mismatches_plt
```

### Supplementary Figure 11: Proportion of genomes with different multi-carbapenemase groups over time

```{r}
#| label: fig-plot-s11
#| fig-cap: Proportion of genomes with different multi-carbapenemase groups over time.. 
#| fig-height: 4
#| warning: false

year <- meta |> 
  mutate(carb_group_plotting = case_when(
    str_count(carb_group, ",") > 1 ~ "other",
    carb_group == "NDM, VIM" ~ "other",
    carb_group == "IMP, OXA" ~ "other",
    carb_group == "IMP, VIM" ~ "other",
    TRUE ~ carb_group )) |> 
  mutate(carb_group_plotting = factor(carb_group_plotting, levels = c("IMP", "IMP, KPC", "IMP, NDM", "KPC", "KPC, NDM", "KPC, OXA", "KPC, VIM", "NDM", "NDM, OXA", "OXA", "OXA, VIM", "VIM", "other"))) |>  
  group_by(year) |> 
  count(carb_group_plotting) |> 
  left_join(meta |> count(year, name = "year_total"), by = "year") |> 
  mutate(carb_group_plotting = str_replace(carb_group_plotting, "OXA", "OXA-48-like")) |> 
  mutate(prop = n/year_total) |> 
  ggplot(aes(x = year, y = prop, fill = carb_group_plotting)) +
  geom_bar_pattern(stat="identity", position = "stack", aes(pattern_fill = carb_group_plotting), pattern_colour = NA, pattern_density = 0.5, pattern_spacing = 0.02) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90)) +
  labs(x = "Year", fill = "Carbapenemase\ngroup", y = "Proportion", pattern_fill = "Carbapenemase\ngroup") +
  scale_fill_manual(values = c("#D55E00", "#D55E00", "#D55E00", "#56B4E9", "#56B4E9", "#56B4E9", "#56B4E9", "#E69F00", "#E69F00", "#CC79A7", "#CC79A7", "#0072B2", "grey")) +
  scale_pattern_fill_manual(values = c("#D55E00", "#56B4E9", "#E69F00", "#56B4E9", "#E69F00","#CC79A7", "#0072B2", "#E69F00", "#CC79A7", "#CC79A7", "#0072B2","#0072B2", "grey"));year
```

### Supplementary Figure 12: Multi-carbapenemase Enterobacterales genomes from NCBI GenBank with fewer than 40 contigs ordered by number of contigs (n=2033)

```{r}
#| label: fig-plot-s12
#| fig-cap: Multi-carbapenemase Enterobacterales genomes from NCBI GenBank with fewer than 40 contigs ordered by number of contigs (n=2033). 
#| fig-height: 4
#| warning: false

ncbi_multicarb_accessions <- carbs_ncbi |> 
  count(accession)  |> 
  filter(n > 1) |> 
  pull(accession)

meta_below40 <- meta_ncbi |> 
  filter(accession %in% ncbi_multicarb_accessions) |> 
  mutate(long_read_platform = ifelse(str_detect(platform, "Oxford|Nanopore|nanopore|OXFORD|pacbio|PacBio|Pacbio|PacBIO|Pacific Bio|GridION|MinION|minION|Hybrid|ONT") & !is.na(platform), "yes", "no")) 

p1 <- meta_below40 |> 
  ggplot(aes(y = assembly_num_contigs, x = reorder(accession, assembly_num_contigs))) +
  geom_point() +
  theme_bw() +
  theme(panel.grid.major.x = element_blank()) +
  geom_hline(yintercept = 20) +
  labs(x = "Genomes", y = "Number of contigs") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank())

p2 <- meta_below40 |> 
  ggplot(aes(x = assembly_num_contigs, fill = long_read_platform)) +
  geom_vline(xintercept = 20) +
  geom_bar(stat = "count") +
  labs(x = "Number of contigs", y = "Number of genomes", fill = "Sequenced with\nlong read platform?") +
  scale_fill_manual(values = c("grey70","grey30")) +
  theme_bw()

p1 + p2 + plot_annotation(tag_levels = 'A')
```

# Statistical analyses

### Pearson's Chi-squared: carbapenemase combinations vs genus

```{r}
#| label: stats-person-genus
#| warning: false

meta_carb <- meta  |> 
  #filter(dereplicate == "unique") |> 
  mutate(carb = ifelse(carb %in% (meta |> filter(dereplicate == "unique") |> count(carb) |> filter(n > 9) |> pull(carb)), carb, "other")) |> 
  count(carb, genus)  |> 
  ungroup() |> 
  arrange(desc(n))  

meta_carb2 <- meta_carb |> 
  mutate(genus = factor(genus)) |> 
  pivot_wider(id_cols = genus, names_from = carb, values_from = n, values_fill = 0) |> 
  column_to_rownames("genus")
sim_test <- chisq.test(meta_carb2, simulate.p.value=TRUE, B=20000);sim_test
residuals_table <- sim_test$stdres
raw_p_values <- 2 * (1 - pnorm(abs(residuals_table)))

# Apply Benjamini-Hochberg (BH/FDR) adjustment across all 300 cells
adjusted_p_values <- p.adjust(raw_p_values, method = "BH")

# Re-shape the adjusted p-values back into the original 12x25 matrix structure
adj_p_matrix <- matrix(adjusted_p_values, nrow = 12, ncol = 25, 
                       dimnames = list(rownames(meta_carb2), colnames(meta_carb2)))

# Find cells where the BH-adjusted p-value is less than 0.05
significant_indices <- which(adj_p_matrix < 0.05, arr.ind = TRUE)

# Display the significant drivers
if(nrow(significant_indices) > 0) {
  results_summary <- data.frame(
    Row = rownames(adj_p_matrix)[significant_indices[, 1]],
    Column = colnames(adj_p_matrix)[significant_indices[, 2]],
    Residual = residuals_table[significant_indices],
    BH_Adj_P = adj_p_matrix[significant_indices]
  )
  print("--- SIGNIFICANT DRIVING CELLS FOUND (BH ADJ) ---")
  print(results_summary[order(results_summary$BH_Adj_P), ]) # Ordered by significance
} else {
  print("No cells were significant after Benjamini-Hochberg adjustment.")
}
```

### Fisher's exact test: genomic location

```{r}
#| label: stats-fishers-genomic-location
#| warning: false

test <- meta |> 
  count(multicarb_type, multicarb_location) |> 
  pivot_wider(id_cols = multicarb_type, names_from = multicarb_location, values_from = n, values_fill = 0) |> 
  column_to_rownames("multicarb_type")
 
fisher.test(test, simulate.p.value = TRUE, B=20000)
chisq.test(test, simulate.p.value = TRUE, B=20000)$stdres
```

### Firth's penalised binary logistic regression: consolidation of carbapenemases onto a single plasmid backbone

```{r}
#| label: stats-firth-consolidation-oneplas
#| warning: false

location_stats <- meta |> 
  #filter(dereplicate == "unique") |> 
  filter(!is.na(year)) |> 
  mutate(year = as.numeric(year)) |> 
  mutate(multicarb_location = droplevels(as.factor(multicarb_location))) |> 
  mutate(multicarb_location = relevel(multicarb_location, ref = "two plas"))

# Create the binary outcome variable 
location_stats$target_binary <- ifelse(
  location_stats$multicarb_location == "one plas", 1, 0
)

# Center the year variable (starts at 1 instead of calendar year) - this brings the intercept down from the billions to a normal scale
location_stats$year_centered <- location_stats$year - min(location_stats$year) + 1

# Run Firth's Penalized Binary Logistic Regression
binary_firth_model <- logistf(
  target_binary ~ year_centered, 
  data = location_stats
)

summary(binary_firth_model)

# Extract the raw log-odds coefficients and profile confidence intervals
log_odds_coef <- coef(binary_firth_model)
log_odds_ci   <- confint(binary_firth_model)

# Convert them to the odds ratio scale 
odds_ratios <- exp(log_odds_coef)
ci_exponentiated <- exp(log_odds_ci)

# print results
results_table <- cbind(OR = odds_ratios, 
                       Lower_95_CI = ci_exponentiated[,1], 
                       Upper_95_CI = ci_exponentiated[,2])

print(results_table["year_centered", , drop = FALSE])
```

### Permutation test: pairwise carbapenemase co-occurence

```{r}
#| label: stats-permutation-cooccurence
#| warning: false

transactions <- meta |> 
  #filter(dereplicate == "unique") |> 
  filter(accession != "GCA_023982975.1") |> # has multiple VIM-63 copies that are skewing results, only isolate with this gene
  mutate(transaction_id = 1:n()) |>
  separate_longer_delim(carb, delim = ", ") |> 
  select(transaction_id, item=carb)

df <- transactions |> 
  mutate(transaction_id = as.factor(transaction_id), item = as.factor(item))

transactions <- levels(df$transaction_id)
items <- levels(df$item)

n_transactions <- length(transactions)
n_items <- length(items)

# Integer indices
transaction_idx <- as.integer(df$transaction_id)
item_idx <- as.integer(df$item)

# spare transaction x item matrix
X <- Matrix::sparseMatrix(
  i = transaction_idx,
  j = item_idx,
  x = 1,
  dims = c(n_transactions, n_items),
  dimnames = list(transactions, items)
)

# function to calculate pair counts
get_pairs_matrix <- function(X) {

  # Cross-products
  cp <- as.matrix(crossprod(X))

  # Self-pairs: choose(count, 2)
  self_counts <- colSums(X * (X - 1) / 2)

  # Replace diagonal of crossprod with correct self-pair counts
  diag(cp) <- self_counts

  cp
}

# calculate observed pair count matrix 
observed <- get_pairs_matrix(X)

# permutation function 
permute_once <- function() {

  permuted_items <- sample(item_idx)

  Xperm <- Matrix::sparseMatrix(
    i = transaction_idx,
    j = permuted_items,
    x = 1,
    dims = c(n_transactions, n_items)
  )

  get_pairs_matrix(Xperm)
}

# assign null quantities 
set.seed(123)
B <- 10000
sum_null <- matrix(
  0,
  nrow = n_items,
  ncol = n_items
)
sumsq_null <- matrix(
  0,
  nrow = n_items,
  ncol = n_items
)
n_ge <- matrix(
  0L,
  nrow = n_items,
  ncol = n_items
)

for (b in seq_len(B)) {

  perm <- permute_once()

  # Null expectation
  sum_null <- sum_null + perm

  # For null SD
  sumsq_null <- sumsq_null + perm^2

  # Number of permutations >= observed
  n_ge <- n_ge + (perm >= observed)

  if (b %% 1000 == 0) {
    message("Permutation ", b, " / ", B)
  }
}

# compute null statistic
expected <- sum_null / B

null_variance <- (
  sumsq_null -
    (sum_null^2 / B)
) / (B - 1)

null_sd <- sqrt(
  pmax(null_variance, 0)
)

p_value <- (n_ge + 1) / (B + 1)

# keep only upper triangle 
idx <- which(
  upper.tri(observed, diag = TRUE),
  arr.ind = TRUE
)

results <- data.frame(
  item_1 = items[idx[, 1]],
  item_2 = items[idx[, 2]],
  observed = observed[idx],
  expected = expected[idx],
  null_sd = null_sd[idx],
  p_value = p_value[idx]
)

# add enrichment and multiple-testing correction
results <- results %>%
  mutate(
    enrichment = ifelse(
      expected > 0,
      observed / expected,
      NA_real_),

    log2_enrichment =
      log2(
        (observed + 0.5) /
        (expected + 0.5)),

    q_value = p.adjust(
      p_value,
      method = "BH")
  ) |> 
  arrange(
    q_value,
    desc(enrichment)
  )

# filtering for pairs that appear in more than 5 genomes
significant_pairs <- results |> 
  filter(q_value < 0.05, enrichment > 1, observed >= 5)

results |> 
  filter(observed > 5) |> 
  select(carb1 = item_1, carb2 = item_2, observed, expected, enrichment, log2_enrichment, p_value, q_value) |> 
  arrange(q_value, desc(enrichment))
```

### Kruskal-Wallis with post-hoc Dunn's and BH correction: AMR genes, AMR gene classes, replicons, length, and toxin-antitoxin systems on plasmids with zero, one, or two or more carbapenemases

```{r}
#| label: stats-kruskal-amr
#| warning: false

multicarb_plas_list <- gene |> 
  filter(location == "plasmid")  |> 
  count(contig_id) |> 
  filter(n > 1) |> 
  pull(contig_id)

singlecarb_plas_list <- gene |> 
  filter(location == "plasmid")  |> 
  count(contig_id) |> 
  filter(n == 1) |> 
  pull(contig_id)

amr_nocarbs <- amr |> 
  filter(!str_detect(gene, "blaNDM|blaKPC|blaOXA-48|blaOXA-181|blaOXA-232|blaOXA-244|blaOXA-204|blaOXA-1205|blaOXA-1207|blaVIM|blaIMP") & !str_detect(closest_reference, "OXA-48|OXA-181|OXA-232|OXA-244|OXA-204|OXA-1205|OXA-1207")) 

amr_per_plas <- contigs |> 
  filter(location == "plasmid") |> 
  mutate(num_replicons = case_when(
    rep_type == "-" ~ 0,
    str_detect(rep_type, ",") ~ str_count(rep_type, ",") +1,
    TRUE ~ 1)) |> 
  select(accession, contig_id, length, num_replicons) |> 
  mutate(num_carb_on_plas = case_when(
    contig_id %in% multicarb_plas_list ~ "Two or\nmore",
    contig_id %in% singlecarb_plas_list ~ "One",
    TRUE ~ "None"
  )) |> 
  select(accession, contig_id, length, num_carb_on_plas, num_replicons) |> 
  distinct() |> 
  left_join(amr_nocarbs, by=c("contig_id", "accession")) |> # excludes carb genes
  group_by(num_carb_on_plas, contig_id, length.x, num_replicons) |> 
  summarize(genes_concat = toString(gene, sep = ", ")) |> 
  mutate(num_genes = case_when(
    genes_concat == "NA" ~ 0,
    str_detect(genes_concat, ",") ~ str_count(genes_concat, ",") +1,
    TRUE ~ 1)) |> 
  left_join(amr_nocarbs, by=c("contig_id")) |> # excludes carb class
  group_by(num_carb_on_plas, contig_id, length.x, num_replicons, num_genes) |>
  separate_longer_delim(subclass, delim = "/") |>
  summarize(num_classes = n_distinct(subclass)) |> 
  mutate(num_classes = ifelse(num_genes == 0, 0, num_classes)) |> 
  mutate(num_carb_on_plas = factor(num_carb_on_plas)) |> 
  ungroup()

ta2 <- ta |> 
  group_by(contig_id) |> 
  summarize(num_TA = n_distinct(id))

amr_per_plas2 <- amr_per_plas |> 
  left_join(ta2, by = "contig_id") |> 
  mutate(num_TA = ifelse(is.na(num_TA), 0, num_TA))  |> 
  filter(num_genes > 0 | num_carb_on_plas == "One" | num_carb_on_plas == "Two or\nmore") # keeping only those with AMR genes

kruskal.test(num_genes ~ num_carb_on_plas, data = amr_per_plas2)
rstatix::dunn_test(num_genes ~ num_carb_on_plas, data = amr_per_plas2, p.adjust.method = "BH")

kruskal.test(num_classes ~ num_carb_on_plas, data = amr_per_plas2)
rstatix::dunn_test(num_classes ~ num_carb_on_plas, data = amr_per_plas2, p.adjust.method = "BH")

kruskal.test(num_replicons ~ num_carb_on_plas, data = amr_per_plas2)
rstatix::dunn_test(num_replicons ~ num_carb_on_plas, data = amr_per_plas2, p.adjust.method = "BH")

kruskal.test(length.x ~ num_carb_on_plas, data = amr_per_plas2)
rstatix::dunn_test(length.x ~ num_carb_on_plas, data = amr_per_plas2, p.adjust.method = "BH")

kruskal.test(num_TA ~ num_carb_on_plas, data = amr_per_plas2)
rstatix::dunn_test(num_TA ~ num_carb_on_plas, data = amr_per_plas2, p.adjust.method = "BH")
```

### Pearson's Chi-squared: genomic location in hypervirulent *K. pneumoniae*

```{r}
#| label: stats-pearson-hvkp
#| warning: false

hvkp_stats <- hvkp_meta |> 
  left_join(multicarb_location_hvkp, by = "accession") |> 
  select(accession, multicarb_location.y, multicarb_type) |> 
  mutate(new_location = ifelse(str_detect(multicarb_location.y, "^vir plas"), "vir plas", "other")) |> 
  count(multicarb_type, new_location) |>
  pivot_wider(id_cols = "multicarb_type", names_from = "new_location", values_from = "n", values_fill = 0) |> 
  column_to_rownames("multicarb_type") |> 
  select(`vir plas`, other)
chisq.test(hvkp_stats)
chisq.test(hvkp_stats)$stdres

results <- lapply(seq_len(nrow(hvkp_stats)), function(i) {

  # 2 x 2 table: focal category vs all other categories
  focal <- hvkp_stats[i, ]
  others <- colSums(hvkp_stats[-i, , drop = FALSE])

  test_tab <- rbind(
    category = focal,
    other_categories = others
  )

  ft <- fisher.test(test_tab)

  data.frame(
    category = rownames(hvkp_stats)[i],
    odds_ratio = unname(ft$estimate),
        CI_lower = ft$conf.int[1],
    CI_upper = ft$conf.int[2],
    p_value = ft$p.value
  )
})

results <- do.call(rbind, results)

# Benjamini-Hochberg FDR correction
results$p_adj_BH <- p.adjust(
  results$p_value,
  method = "BH"
)
results$prop_other <- hvkp_stats[, "other"] / sum(hvkp_stats[, "other"])
results$prop_virulence <- hvkp_stats[, "vir plas"] / sum(hvkp_stats[, "vir plas"])
results
```

### Anderson-Darling test for normality/uniform distribution: distance of carbapenemase genes on the same plasmid backbone

```{r}
#| label: stats-anderson-darling
#| warning: false

test <- tn4 |> filter(carb_group == "KPC") |> pull(prop)
goftest::ad.test(test, null = "punif", min = 0, max = 0.500001)

test <- tn4 |> filter(carb_group == "NDM") |> pull(prop)
goftest::ad.test(test, null = "punif", min = 0, max = 0.500001)

test <- tn4 |> filter(carb_group == "OXA") |> pull(prop)
goftest::ad.test(test, null = "punif", min = 0, max = 0.500001)

test <- tn4 |> filter(carb_group == "KPC, NDM") |> pull(prop)
goftest::ad.test(test, null = "punif", min = 0, max = 0.500001)

test <- tn4 |> filter(carb_group == "NDM, OXA") |> pull(prop)
goftest::ad.test(test, null = "punif", min = 0, max = 0.500001)
```

### Exact binomal test: novel carbapenemase alleles

```{r}
#| label: stats-exact-binomial
#| warning: false

mismatches2 <- amr |> 
  filter(accession %in% (meta |> pull(accession))) |> 
  filter(str_detect(gene, "blaNDM|blaKPC|blaOXA-48|blaOXA-181|blaOXA-232|blaOXA-244|blaOXA-204|blaOXA-1205|blaOXA-1207|blaVIM|blaIMP") | str_detect(closest_reference, "OXA-48|OXA-181|OXA-232|OXA-244|OXA-204|OXA-1205|OXA-1207")) |> 
  filter(status != "ALLELEX" & status != "PARTIAL_CONTIG_ENDX") |> 
  left_join(meta_raw |> select(accession, multicarb_type, multicarb_location, carb), by = "accession") |> 
  filter(!is.na(multicarb_type)) |> 
  mutate(multicarb_type = ifelse(multicarb_type == "different types", "different families", "same family")) |> 
  group_by(gene, status) |> 
  count(multicarb_type)

binom.test(x = (mismatches2  |> ungroup()  |> filter(multicarb_type == "same family")  |> summarize(n = sum(n)) |> pull()), 
  n = (mismatches2  |> ungroup()  |> summarize(n = sum(n)) |> pull()), 
  p=0.5, 
  alternative = "two.sided")

```

### Firth's penalised binary logistic regression: longitudinal KPC/NDM

```{r}
#| label: stats-firth-year-kn
#| warning: false

kn_stats <- meta |> 
  #filter(dereplicate == "unique") |> 
  filter(!is.na(year)) |> 
  mutate(year = as.numeric(year)) |> 
  mutate(carb_group_plotting = case_when(
    str_count(carb_group, ",") > 1 ~ "other",
    carb_group == "NDM, VIM" ~ "other",
    carb_group == "IMP, OXA" ~ "other",
    carb_group == "IMP, VIM" ~ "other",
    TRUE ~ carb_group )) |> 
  mutate(carb_group_plotting = droplevels(as.factor(carb_group_plotting))) |> 
  mutate(carb_group_plotting = relevel(carb_group_plotting, ref = "KPC, NDM"))

# Create the binary outcome variable 
kn_stats$target_binary <- ifelse(
  kn_stats$carb_group_plotting == "KPC, NDM", 1, 0
)

# Center the year variable (starts at 1 instead of calendar year) - this brings the intercept down from the billions to a normal scale
kn_stats$year_centered <- kn_stats$year - min(kn_stats$year) + 1

# Run Firth's Penalized Binary Logistic Regression
binary_firth_model <- logistf(
  target_binary ~ year_centered, 
  data = kn_stats
)

summary(binary_firth_model)

# Extract the raw log-odds coefficients and profile confidence intervals
log_odds_coef <- coef(binary_firth_model)
log_odds_ci   <- confint(binary_firth_model)

# Convert them to the odds ratio scale 
odds_ratios <- exp(log_odds_coef)
ci_exponentiated <- exp(log_odds_ci)

# print results
results_table <- cbind(OR = odds_ratios, 
                       Lower_95_CI = ci_exponentiated[,1], 
                       Upper_95_CI = ci_exponentiated[,2])

print(results_table["year_centered", , drop = FALSE])
```

### Firth's penalised binary logistic regression: longitudinal NDM/OXA

```{r}
#| label: stats-firth-year-no
#| warning: false

no_stats <- meta |> 
  #filter(dereplicate == "unique") |> 
  filter(!is.na(year)) |> 
  mutate(year = as.numeric(year)) |> 
  mutate(carb_group_plotting = case_when(
    str_count(carb_group, ",") > 1 ~ "other",
    carb_group == "NDM, VIM" ~ "other",
    carb_group == "IMP, OXA" ~ "other",
    carb_group == "IMP, VIM" ~ "other",
    TRUE ~ carb_group )) |> 
  mutate(carb_group_plotting = droplevels(as.factor(carb_group_plotting))) |> 
  mutate(carb_group_plotting = relevel(carb_group_plotting, ref = "NDM, OXA"))

# Create the binary outcome variable 
no_stats$target_binary <- ifelse(
  no_stats$carb_group_plotting == "NDM, OXA", 1, 0
)

# Center the year variable (starts at 1 instead of calendar year) - this brings the intercept down from the billions to a normal scale
no_stats$year_centered <- no_stats$year - min(no_stats$year) + 1

# Run Firth's Penalized Binary Logistic Regression
binary_firth_model <- logistf(
  target_binary ~ year_centered, 
  data = no_stats
)

summary(binary_firth_model)

# Extract the raw log-odds coefficients and profile confidence intervals
log_odds_coef <- coef(binary_firth_model)
log_odds_ci   <- confint(binary_firth_model)

# Convert them to the odds ratio scale 
odds_ratios <- exp(log_odds_coef)
ci_exponentiated <- exp(log_odds_ci)

# print results
results_table <- cbind(OR = odds_ratios, 
                       Lower_95_CI = ci_exponentiated[,1], 
                       Upper_95_CI = ci_exponentiated[,2])

print(results_table["year_centered", , drop = FALSE])

```

### Firth's penalised binary logistic regression: consolidation of KPC/NDM onto one plasmid backbone

```{r}
#| label: stats-firth-consolidation-kn
#| warning: false

kn_stats2 <- meta |> 
  #filter(dereplicate == "unique") |> 
  filter(!is.na(year)) |> 
  mutate(year = as.numeric(year)) |> 
  filter(carb_group == "KPC, NDM")

kn_oneplas_list <- gene |> 
  filter(location == "plasmid") |> 
  group_by(accession, contig_id) |> 
  summarize(gene = toString(gene, sep=", ")) |> 
  filter(str_detect(gene, "NDM") & str_detect(gene, "KPC")) |> 
  pull(accession) # one duplicate accession

# Create the binary outcome variable 
kn_stats2$target_binary <- ifelse(
  kn_stats2$accession %in% kn_oneplas_list, 1, 0
)

# Center the year variable (starts at 1 instead of calendar year) - this brings the intercept down from the billions to a normal scale
kn_stats2$year_centered <- kn_stats2$year - min(kn_stats2$year) + 1

# Run Firth's Penalized Binary Logistic Regression
binary_firth_model <- logistf(
  target_binary ~ year_centered, 
  data = kn_stats2
)

summary(binary_firth_model)

# Extract the raw log-odds coefficients and profile confidence intervals
log_odds_coef <- coef(binary_firth_model)
log_odds_ci   <- confint(binary_firth_model)

# Convert them to the odds ratio scale 
odds_ratios <- exp(log_odds_coef)
ci_exponentiated <- exp(log_odds_ci)

# print results
results_table <- cbind(OR = odds_ratios, 
                       Lower_95_CI = ci_exponentiated[,1], 
                       Upper_95_CI = ci_exponentiated[,2])

print(results_table["year_centered", , drop = FALSE])
```

### Firth's penalised binary logistic regression: consolidation of NDM/OXA onto one plasmid backbone

```{r}
#| label: stats-firth-consolidation-no
#| warning: false

no_stats2 <- meta |> 
  #filter(dereplicate == "unique") |> 
  filter(!is.na(year)) |> 
  mutate(year = as.numeric(year)) |> 
  filter(carb_group == "NDM, OXA")

no_oneplas_list <- gene |> 
  filter(location == "plasmid") |> 
  group_by(accession, contig_id) |> 
  summarize(gene = toString(gene, sep=", ")) |> 
  filter(str_detect(gene, "NDM") & str_detect(gene, "OXA")) |> 
  pull(accession)

# 1. Create the binary outcome variable 
no_stats2$target_binary <- ifelse(
  no_stats2$accession %in% no_oneplas_list, 1, 0
)

# Center the year variable (starts at 1 instead of calendar year) - this brings the intercept down from the billions to a normal scale
no_stats2$year_centered <- no_stats2$year - min(no_stats2$year) + 1

# Run Firth's Penalized Binary Logistic Regression
binary_firth_model <- logistf(
  target_binary ~ year_centered, 
  data = no_stats2
)

summary(binary_firth_model)

# Extract the raw log-odds coefficients and profile confidence intervals
log_odds_coef <- coef(binary_firth_model)
log_odds_ci   <- confint(binary_firth_model)

# Convert them to the odds ratio scale 
odds_ratios <- exp(log_odds_coef)
ci_exponentiated <- exp(log_odds_ci)

# print results
results_table <- cbind(OR = odds_ratios, 
                       Lower_95_CI = ci_exponentiated[,1], 
                       Upper_95_CI = ci_exponentiated[,2])

print(results_table["year_centered", , drop = FALSE])
```
